Preliminary analysis of Greenland narwhals trajectories with smoothSDE

### 2025-03-17 R/application/CVM/fit_cvm.Rmd
update application to narwhal data

estimates_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Est")
std_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Std")

@@

We process the results and show a table with the estimates and their standard deviations.

```{r render coefficient estimates table baseline_ctcrw0 }

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_ctcrw0$log_sigma_obs, 
                               std_bas_ctcrw0$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_ctcrw0$coeff_fe, 
                          std_bas_ctcrw0$coeff_fe, 
                          rownames(estimates_bas_ctcrw0$coeff_fe))

# Combine all data frames into one
table_data_baseline_ctcrw0 <- bind_rows(log_sigma_obs, coeff_fe)

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data_baseline_ctcrw0, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
```

From this we can obtain $95\%$ confidence intervals for the parameters $\tau$, $\nu$ and $\sigma_{obs}$.

```{r CI baseline_ctcrw0 parameters}

post_coeff_bas_ctrcw0=baseline_ctcrw0$post_coeff(n_post=10000)
post_par_bas_ctcrw0=
  list("sigma_obs"=1000*exp(post_coeff_bas_ctrcw0$log_sigma_obs),
              "tau"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"tau.(Intercept)"]),
              "nu"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"nu.(Intercept)"]))
mean_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,mean)
sd_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,sd)
quant_par_bas_ctcrw0 <- lapply(post_par_bas_ctcrw0, 
                               quantile, probs = c(0.025, 0.975))

We include the results in a latex table.

```{r CI table baseline_ctcrw0 parameters,results='asis',message=FALSE}

parameter_names <- names(mean_par_bas_ctcrw0)
estimates <- mean_par_bas_ctcrw0
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw0[[param]][1], 
          quant_par_bas_ctcrw0[[param]][2])
})

table_data <- data.frame(
  Parameter = c("$\\sigma_{obs}$","$\\tau$", "$\\nu$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw0),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```{r,results='asis',echo=FALSE}
# Create an xtable object
xtab <- xtable(
  table_data,
  caption = "Baseline estimations",
  label = "table:baseline_estimations"
)

# Customize xtable options
print(
  xtab,
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),
  table.placement="H"
)
```

@@

```{r render estimates table baselines with fixed mesaurement error,results='asis', message=FALSE}

for (i in seq_along(list_baseline_ctcrw0)) {
  estimates=list_baseline_ctcrw0_est[[i]]
  std=list_baseline_ctcrw0_std[[i]]

  # Create data frames for each set
  log_sigma_obs <- combine_to_df(estimates$log_sigma_obs, 
                               std$log_sigma_obs, "log_sigma_obs")
  coeff_fe <- combine_to_df(estimates$coeff_fe, 
                          std$coeff_fe, 
                          rownames(estimates$coeff_fe))

  # Combine all data frames into one
  table_data <- bind_rows(log_sigma_obs, coeff_fe)

  latex_table <- xtable(
    table_data, 
    caption = paste(
      "Parameter Estimates and Standard Errors for fixed \\(\\sigma_{obs}=\\) ",
      sigma_obs[i]," m"),
    xtable.table.placement="H")

  # Print LaTeX table
  print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")

@@

We see a clear effect of the fixed measurement error on the estimates of $\tau$, and a slight influence of $\nu$. We also remark that as the measurement error grows, the uncertainty on the estimates of $\tau$ and $\nu$ increases. Overall, knowing or estimating precisely the measurement error will be crucial in the analysis.

@@

#Fit baseline with measurement error estimated from the data
baseline_ctcrw1<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
                    fixpar=c("mu1","mu2"))
baseline_ctcrw1$fit()

```{r results baseline_ctcrw1}

estimates_bas_ctcrw1=as.list(baseline_ctcrw1$tmb_rep(),what="Est")
std_bas_ctcrw1=as.list(baseline_ctcrw1$tmb_rep(),what="Std")

```{r render estimates table baseline_ctcrw1}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_ctcrw1$log_sigma_obs,
                               std_bas_ctcrw1$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_ctcrw1$coeff_fe,
                          std_bas_ctcrw1$coeff_fe,
                          rownames(estimates_bas_ctcrw1$coeff_fe))
log_lambda <- combine_to_df(estimates_bas_ctcrw1$log_lambda, 
                            std_bas_ctcrw1$log_lambda, 
                            rownames(estimates_bas_ctcrw1$log_lambda))
coeff_re <- combine_to_df(estimates_bas_ctcrw1$coeff_re, 
                          std_bas_ctcrw1$coeff_re, 
                          rownames(estimates_bas_ctcrw1$coeff_re))

# Combine all data frames into one
table_data_baseline_ctcrw1 <- bind_rows(log_sigma_obs, coeff_fe, log_lambda, coeff_re)

```{r, results='asis', echo=FALSE}
latex_table <- xtable(table_data_baseline_ctcrw1, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
```

As previously, we can obtain $95\%$ confidence intervals for the parameters $\tau$, $\nu$ and $\sigma_{obs}$, and include the results in a latex table.

```{r CI baseline_ctcrw1 parameters}

post_coeff_bas_ctrcw1=baseline_ctcrw1$post_coeff(n_post=10000)
post_par_bas_ctcrw1=list(
  "sigma_obs"=1000*exp(post_coeff_bas_ctrcw1$log_sigma_obs),
  "tau"=exp(post_coeff_bas_ctrcw1$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_ctrcw1$coeff_fe[,"nu.(Intercept)"]),
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_ctrcw1$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_ctrcw1$log_lambda[,"nu.s(ID)"])))

mean_par_bas_ctcrw1=lapply(post_par_bas_ctcrw1,mean)
sd_par_bas_ctcrw1=lapply(post_par_bas_ctcrw1,sd)
quant_par_bas_ctcrw1 <- lapply(post_par_bas_ctcrw1, 
                               quantile, probs = c(0.025, 0.975))

```{r CI table baseline_ctcrw1 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_ctcrw1)
estimates <- mean_par_bas_ctcrw1
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw1[[param]][1], 
          quant_par_bas_ctcrw1[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\sigma_{obs}$","$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw1),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```
```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "Baseline estimations with random effects",
  label = "table:baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```

@@

### RACVM without random individual effects

We can also test whether adding a constant rotational $\omega$ in the SDE model would make sense. For this, we fit a RACVM model.

```{r fit baseline racvm,message=FALSE}

#initial parameters
par0 <- c(0,0,1,4,0)

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~1,
                 nu=~1,
                 omega=~1)

#Fit baseline with measurement error estimated from the data
baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM",
                          response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
                    fixpar=c("mu1","mu2"))
baseline_racvm0$fit()

```{r results baseline_racvm0}

estimates_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Est")
std_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Std")

```{r render estimates table baseline_racvm0}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_racvm0$log_sigma_obs, 
                               std_bas_racvm0$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_racvm0$coeff_fe,
                          std_bas_racvm0$coeff_fe, 
                          rownames(estimates_bas_racvm0$coeff_fe))

# Combine all data frames into one
table_data_baseline_racvm0 <- bind_rows(log_sigma_obs, coeff_fe)

```{r, results='asis', echo=FALSE}
latex_table <- xtable(table_data_baseline_racvm0, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,table.placement="H")
```

The estimate of $\omega$ is $0$. This indicates that a \textbf{constant} parameter $\omega$ is irrelevant in the model. As a consequence, the final baseline model we keep is \texttt{baseline\_ctcrw1} which corresponds to a CTCRW with constant parameters $\tau$ and $\nu$ and mean velocity fixed to zero.

@@

#Fit baseline with measurement error estimated from the data
baseline_ctcrw1_24h<- SDE$new(
  formulas = formulas,data = dataBE24,type = "CTCRW",
  response = c("x","y"),par0 = par0,
  other_data=list("log_sigma_obs0"=log(0.05)),
  fixpar=c("mu1","mu2"))

@@

```{r results baseline_ctcrw1_24h}

estimates_bas_ctcrw1_24h=as.list(baseline_ctcrw1_24h$tmb_rep(),what="Est")
std_bas_ctcrw1_24h=as.list(baseline_ctcrw1_24h$tmb_rep(),what="Std")

```{r render estimates table baseline_ctcrw1_24h}

# Create data frames for each set
log_sigma_obs <- combine_to_df(
  estimates_bas_ctcrw1_24h$log_sigma_obs, 
  std_bas_ctcrw1_24h$log_sigma_obs, "log_sigma_obs")

coeff_fe <- combine_to_df(estimates_bas_ctcrw1_24h$coeff_fe, 
                          std_bas_ctcrw1_24h$coeff_fe, 
                          rownames(estimates_bas_ctcrw1_24h$coeff_fe))

# Combine all data frames into one
table_data_baseline_ctcrw1_24h <- bind_rows(log_sigma_obs, coeff_fe)

```{r, results='asis',echo=FALSE}
latex_table <- xtable(table_data_baseline_ctcrw1_24h, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,
      table.placement="H")
```

The standard deviations are wider due to the smaller amount of data, but also due to the fact that the measurement error is estimated higher than before. $\tau$ seems to be estimated slightly below the value we obtained for the dataset \texttt{dataBE12}, but the difference seems reasonable as the CI of the two estimates overlap.

```{r fit baseline sensitivity analysis tagging effect fixed error,message=FALSE}

sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))

#Fit baseline with measurement error 
baseline_ctcrw1_24h_fixed_error<- SDE$new(formulas = formulas,
                                          data = dataBE24,type = "CTCRW",
                                          response = c("x","y"),
                                          par0 = par0,
                                          other_data=list("H"=H),
                                          fixpar=c("mu1","mu2"))
baseline_ctcrw1_24h_fixed_error$fit()

```{r results baseline_ctcrw1_24h_fixed_error}

estimates_bas_ctcrw1_24h_fixed_error=as.list(
  baseline_ctcrw1_24h_fixed_error$tmb_rep(),what="Est")
std_bas_ctcrw1_24h_fixed_error=as.list(
  baseline_ctcrw1_24h_fixed_error$tmb_rep(),what="Std")

```{r render estimates table baseline_ctcrw1_24h_fixed_error }

# Create data frames for each set
log_sigma_obs <- combine_to_df(
  estimates_bas_ctcrw1_24h_fixed_error$log_sigma_obs,
  std_bas_ctcrw1_24h_fixed_error$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(
  estimates_bas_ctcrw1_24h_fixed_error$coeff_fe, 
  std_bas_ctcrw1_24h_fixed_error$coeff_fe, 
  rownames(estimates_bas_ctcrw1_24h_fixed_error$coeff_fe))

# Combine all data frames into one
table_data_baseline_ctcrw1_24h_fixed_error <- bind_rows(log_sigma_obs, coeff_fe)

```{r, results='asis',echo=FALSE}
latex_table <- xtable(table_data_baseline_ctcrw1_24h_fixed_error, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,
      table.placement="H")
```

When the measurement error is fixed to the value we initially estimated from the whole dataset \texttt{dateBE12}, we observe a big decrease in the estimate of $\tau$. This indeed suggest a shift in the behavior of the whales within the time interval [12h,24h] after tagging, which might invalidate our results.

@@

sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))

@@

```{r results response_ctcrw1}

estimates_res_ctcrw1=as.list(response_ctcrw1$tmb_rep(),what="Est")
std_res_ctcrw1=as.list(response_ctcrw1$tmb_rep(),what="Std")

```{r render estimates table response_ctcrw1}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_res_ctcrw1$log_sigma_obs, 
                               std_res_ctcrw1$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_res_ctcrw1$coeff_fe,
                          std_res_ctcrw1$coeff_fe, 
                          rownames(estimates_res_ctcrw1$coeff_fe))

# Combine all data frames into one
table_data_response_ctcrw1 <- bind_rows(log_sigma_obs, coeff_fe)

```{r, results='asis', echo=FALSE}
latex_table <- xtable(table_data_response_ctcrw1, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,
      table.placement="H")
```

The estimated standard deviation of the random effects can be computed with \texttt{sdev} method in \texttt{smoothSDE}.

```{r}
response_ctcrw1$sdev()
```

The estimates are lower than those obtained from the data before exposure. This may happen because we have short trajectories before exposure, that may not represent the whole range of possible movement for each narwhal, and therefore there may be more varieties of movements between the different individuals. On the contrary, the trajectories during exposure are longer so we can reasonably say that they represent better the range of possible movement. The lower standard deviation of the random effects might then indicate that on the long run, the narwhals in the study tend to have similar movement characteristics.

@@

The intercept of the parameters $\tau$ and $\nu$ are considered as offset in the statistical model, and fixed using the estimates we get from the baseline model \texttt{baseline\_ctcrw1}. The measurement error is an offset 7that is fixed to the value that has been estimated from the data before exposure.

@@

# Fit response with measurement error fixed to baseline value
sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))
response_ctcrw_sp<- SDE$new(formulas = formulas,data = dataAE,type = "CTCRW",
                            response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

# fix random effects to baseline values
response_ctcrw_sp$update_coeff_re(
  c(rep(0,9),baseline_ctcrw1$coeff_re()[1:6,1],rep(0,9)
    ,baseline_ctcrw1$coeff_re()[7:12,1]))
response_ctcrw_sp$update_lambda(
  c(1,baseline_ctcrw1$lambda()[1,1],1,baseline_ctcrw1$lambda()[2,1]))
response_ctcrw_sp$update_coeff_fe(baseline_ctcrw1$coeff_fe())
#use the indices of the variable in the map for TMB 
#to keep the coefficients in the right order
#otherwise we may have issues in function post_coeff
response_ctcrw_sp$update_map(list(coeff_re=factor(
  c(1:9,rep(NA,6),10:18,rep(NA,6))),
  coeff_fe=factor(rep(NA,4)),
  log_lambda=factor(c(1,NA,2,NA))))

@@

```{r results response spline}

estimates_res_ctcrw_sp=as.list(response_ctcrw_sp$tmb_rep(),what="Est")
std_res_ctcrw_sp=as.list(response_ctcrw_sp$tmb_rep(),what="Std")

```{r render estimates table response_ctcrw_sp}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_res_ctcrw_sp$log_sigma_obs, 
                               std_res_ctcrw_sp$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_res_ctcrw_sp$coeff_fe,
                          std_res_ctcrw_sp$coeff_fe,
                          rownames(estimates_res_ctcrw_sp$coeff_fe))
coeff_re <- combine_to_df(estimates_res_ctcrw_sp$coeff_re, 
                          std_res_ctcrw_sp$coeff_re, 
                          rownames(estimates_res_ctcrw_sp$coeff_re))
log_lambda <- combine_to_df(estimates_res_ctcrw_sp$log_lambda, 
                            std_res_ctcrw_sp$log_lambda, 
                            rownames(estimates_res_ctcrw_sp$log_lambda))

# Combine all data frames into one
table_data_response_ctcrw_sp <- bind_rows(
  log_sigma_obs, coeff_fe,coeff_re,log_lambda)

```{r,results='asis',echo=FALSE}
latex_table <- xtable(table_data_response_ctcrw_sp, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,
      table.placement="H")
```

@@

The estimated smooth functions depend on the number of knots we choose in the splines. For $\nu$, the estimated smooth does not change as we increase gradually the degree of freedom to $10$, meaning that only a few knots (1 or 2) are needed. The effect of the presence of the ship on the parameter $\nu$ is easily interpreted : the closer the narwhals are to the ship, the higher their velocity, likely to move away or avoid the ship. Now, the estimated smooth function for $\tau$ requires more knots and is harder to interpret. Though there seem to be a decrease in persistence close to the ship, at a medium distance around $20$ km the persistence of the movement seem to increase. This is likely connected to figure 7 in [Heide-Jorgensen], where depending on how the narwhals were first exposed to the ship, they either change direction (lower persistence) or persist in their heading to flee away (higher persistence).

@@

# Fit response with measurement error fixed to baseline value
sigma_obs=exp(estimates_bas_ctcrw1$log_sigma_obs)
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))

@@

```{r results response_ctcrw_exp}

estimates_res_ctcrw_exp=as.list(response_ctcrw_exp$tmb_rep(),what="Est")
std_res_ctcrw_exp=as.list(response_ctcrw_exp$tmb_rep(),what="Std")

```{r render estimates table response_ctcrw_exp}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_res_ctcrw_exp$log_sigma_obs, 
                               std_res_ctcrw_exp$log_sigma_obs, 
                               "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_res_ctcrw_exp$coeff_fe, 
                          std_res_ctcrw_exp$coeff_fe,
                          rownames(estimates_res_ctcrw_exp$coeff_fe))

# Combine all data frames into one
table_data_response_ctcrw_exp <- bind_rows(log_sigma_obs, coeff_fe)

```{r, results='asis', echo=FALSE}
latex_table <- xtable(table_data_response_ctcrw_exp,
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE,
      table.placement="H")
```

We can also compute $95 \%$ CI for the coefficients in the log-linear model.

```{r CI response_ctcrw_exp parameters}

post_coeff_res_exp=response_ctcrw_exp$post_coeff(n_post=10000)
post_par_res_exp=list(
  "alpha_tau"=post_coeff_res_exp$coeff_fe[,"tau.ExpShip"],
  "alpha_nu"=post_coeff_res_exp$coeff_fe[,"nu.ExpShip"])

mean_par_res_exp=lapply(post_par_res_exp,mean)
sd_par_res_exp=lapply(post_par_res_exp,sd)
quant_par_res_exp <- lapply(post_par_res_exp, 
                               quantile, probs = c(0.025, 0.975))

```{r CI table response_ctcrw_exp parameters, message=FALSE}

parameter_names <- names(mean_par_res_exp)
estimates <- mean_par_res_exp
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_res_exp[[param]][1], 
          quant_par_res_exp[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\alpha_{\\tau}$","$\\alpha_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_res_exp),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "Response log linear estimations",
  label = "table:response_log_linear_estimations"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),
  table.placement="H"
)
```

@@

The results are indeed similar to what we obtain using cubic splines with a low degree of freedom.

@@

post_coeff_re=post_coeff$coeff_re
post_coeff_fe=post_coeff$coeff_fe
post_coeff_lambda=exp(post_coeff$log_lambda)
post_coeff_sigma_obs=exp(post_coeff$log_sigma_obs)

@@

    sigma_obs=post_coeff_sigma_obs[i,1]
    H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_post))

@@

```
```{r propagated CI alpha log linear coefficients, message=FALSE}

@@

@@

We know want to investigate how the fitted model match the actual observations. To fo this, we simulate trajectories from the fitted model before and during exposure. First, we ignore that the narwhals move in a restricted area and only look at the shape of the trajectories.

>>>>

get_raw_latex_tab<- function(sde,show=TRUE) {

  estimates=as.list(sde$tmb_rep(),what="Est")
  std=as.list(sde$tmb_rep(),what="Std")

  # Create data frames for each set
  log_sigma_obs <- combine_to_df(estimates$log_sigma_obs, 
                                 std$log_sigma_obs, "log_sigma_obs")
  coeff_fe <- combine_to_df(estimates$coeff_fe,
                            std$coeff_fe, 
                            rownames(estimates$coeff_fe))
  if (!(is.null(rownames(estimates$coeff_re)))) {

    coeff_re <- combine_to_df(estimates$coeff_re,
                            std$coeff_re, 
                            rownames(estimates$coeff_re))

    log_lambda <- combine_to_df(estimates$log_lambda,
                            std$log_lambda, 
                            rownames(estimates$log_lambda))

     # Combine all data frames into one
    table_data <- bind_rows(log_sigma_obs, coeff_fe,coeff_re,log_lambda)
  } else {
    table_data <- bind_rows(log_sigma_obs, coeff_fe)
  }

  latex_table <- xtable(table_data, 
                        caption = "Parameter Estimates and Standard Errors")

  if (show) {
    print(
    latex_table,type="latex",
    include.rownames = FALSE,
    sanitize.text.function = identity,
    hline.after = c(-1, 0, nrow(table_data)-1),  
    table.placement="H")
  }

  return (latex_table)

raw_ctcrw0_tab=get_raw_latex_tab(baseline_ctcrw0)

@@

```{r}

get_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,show=FALSE) {

  post_coeff=sde$post_coeff(n_post=100000)
  if (ncol(post_coeff$log_lambda)==0) {

  post_par=list(
  "tau"=exp(post_coeff$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff$coeff_fe[,"nu.(Intercept)"]))

  parameter_names=c("$\\tau$", "$\\nu$")

  } else {
  post_par=list(
  "tau"=exp(post_coeff$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff$coeff_fe[,"nu.(Intercept)"]),
  "sigma_tau"=1/sqrt(exp(post_coeff$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff$log_lambda[,"nu.s(ID)"])))

  parameter_names=c("$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$")
  }

  if (include_sigma_obs) {
    post_par$sigma_obs=1000*exp(post_coeff$log_sigma_obs)
  }

  mean_par=lapply(post_par,mean)
  sd_par=lapply(post_par,sd)
  quant_par <- lapply(post_par,
                      quantile, probs = c(0.025, 0.975))

  parameter_names <- names(mean_par)
  estimates <- mean_par
  conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
          quant_par[[param]][2])
    })

  if (include_sigma_obs) {
    parameter_names=c(parameter_names,"$\\sigma_{\\obs}$")
  }
  table_data <- data.frame(
    Parameter = parameter_names,
  Estimate = sprintf("$%.2f$", mean_par),
  CI = conf_intervals,
  stringsAsFactors = FALSE)

  xtab <- xtable(
  table_data,
  caption = paste(sde$type(),"baseline estimations"),
  label = paste0("table:",sde$type(),"_baseline_estimations"))

  if (show) {
  print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H")
  }

  return(xtab)

final_ctcrw0_tab=get_final_baseline_tab(baseline_ctcrw0,show=TRUE)
```

@@

@@

We see a clear effect of the fixed measurement error on the estimates of $\tau$, and a slight influence of $\nu$. We also remark that as the measurement error grows, the uncertainty on the estimates of $\tau$ and $\nu$ incre.ases. Overall, knowing or estimating precisely the measurement error will be crucial in the analysis. On the whole data, it seems more likely to have an error about 50 metres. We now fix the measurement error to 50m

@@

H=array(rep(0.05^2*diag(2),n_pre12),dim=c(2,2,n_pre12))

#Fit baseline with measurement error estimated from the data
baseline_ctcrw1<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))
baseline_ctcrw1$fit()

```
```{r}
final_tab_ctcrw1=get_final_baseline_tab(baseline_ctcrw1,show=TRUE)
```

@@

@@

#Fit baseline with measurement error estimated from the data
baseline_ctcrw1_24h<- SDE$new(
  formulas = formulas,data = dataBE24,type = "CTCRW",
  response = c("x","y"),par0 = par0,
  other_data=list("H"=H),
  fixpar=c("mu1","mu2"))

baseline_ctcrw1_24h$update_lambda(c(exp(3),exp(5)))
baseline_ctcrw1_24h$update_map(list("log_lambda"=factor(c(NA,NA))))

@@

```
```{r}
final_tab_ctcrw1_24h=get_final_baseline_tab(baseline_ctcrw1_24h,show=TRUE)
```

The standard deviations are wider due to the smaller amount of data, but also due to the fact that the measurement error is estimated higher than before. $\tau$ seems to be estimated below the value we obtained for the dataset \texttt{dataBE12}, but the CIs of the two estimates overlap.

@@

H=array(rep(0.05^2*diag(2),n_post),dim=c(2,2,n_post))

@@

```{r}
final_tab_response_ctcrw1=get_final_baseline_tab(response_ctcrw1,show=TRUE)
```

The estimates are lower than those obtained from the data before exposure with first 12h removed, but quite similar to the data before exposure with first 24h removed.

@@

The intercept of the parameters $\tau$ and $\nu$ are considered as offset in the statistical model, and fixed using the estimates we get from the baseline model \texttt{baseline\_ctcrw1}. The measurement error is an offset that is fixed to the value that has been estimated from the data before exposure.

@@

# Fit response 
H=array(rep(0.05^2*diag(2),n_post),dim=c(2,2,n_post))
response_ctcrw_sp<- SDE$new(formulas = formulas,data = dataAE,type = "CTCRW",
                            response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

# fix random effects to baseline values
response_ctcrw_sp$update_coeff_re(
  c(rep(0,9),baseline_ctcrw1$coeff_re()[1:6,1],rep(0,9)
    ,baseline_ctcrw1$coeff_re()[7:12,1]))
response_ctcrw_sp$update_lambda(
  c(1,baseline_ctcrw1$lambda()[1,1],1,baseline_ctcrw1$lambda()[2,1]))
response_ctcrw_sp$update_coeff_fe(baseline_ctcrw1$coeff_fe())
#use the indices of the variable in the map for TMB 
#to keep the coefficients in the right order
#otherwise we may have trouble in function post_coeff
response_ctcrw_sp$update_map(list(coeff_re=factor(
  c(1:9,rep(NA,6),10:18,rep(NA,6))),
  coeff_fe=factor(rep(NA,4)),
  log_lambda=factor(c(1,NA,2,NA))))

@@

@@

The estimated smooth functions depend on the number of knots we choose in the splines. For $\nu$, the estimated smooth does not change as we increase gradually the degree of freedom to $10$, meaning that only a few knots (1 or 2) are needed. The effect of the presence of the ship on the parameter $\nu$ is easily interpreted : the closer the narwhals are to the ship, the higher their velocity, likely to move away or avoid the ship. Now, the estimated smooth function for $\tau$ requires more knots and is harder to interpret. Though there seem to be a decrease in persistence close to the ship, at a medium distance around $20$ km the persistence of the movement seem to increase. This is likely connected to figure 7 in [Heide-Jorgensen], where depending on how the narwhals were first exposed to the ship, they either change direction (lower persistence) or persist in their heading to flee away (higher persistence). This may be also linked to the complex fjords boundaries in which the narwhals move that can trigger very different reaction depending on where the narwhals has been exposed to the ship.

@@

# Fit response with measurement error fixed to baseline value
H=array(rep(0.05^2*diag(2),n_post),dim=c(2,2,n_post))

@@

```{r}
get_final_response_tab<-function(sde,show=TRUE) {

  post_coeff=sde$post_coeff(n_post=10000)
  post_par=list(
  "alpha_tau"=post_coeff$coeff_fe[,"tau.ExpShip"],
  "alpha_nu"=post_coeff$coeff_fe[,"nu.ExpShip"])

  mean_par=lapply(post_par,mean)
  sd_par=lapply(post_par,sd)
  quant_par <- lapply(post_par,
                      quantile, probs = c(0.025, 0.975))

  parameter_names <- names(mean_par)
  estimates <- mean_par
  conf_intervals <- sapply(parameter_names, function(param) {
    sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
            quant_par[[param]][2])
  })

  table_data <- data.frame(
    Parameter = 
      c("$\\alpha_{\\tau}$","$\\alpha_{\\nu}$"),
    Estimate = sprintf("$%.2f$", mean_par),
    CI = conf_intervals,
    stringsAsFactors = FALSE
  )
  xtab <- xtable(
    table_data,
    caption = "Response log linear estimations",
    label = "table:response_log_linear_estimations"
  )

  if (show) {
    print(
    xtab,type="latex",
    include.rownames = FALSE,
    sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
    hline.after = c(-1, 0, nrow(table_data)),
    table.placement="H")
  }

  return (xtab)

final_tab_response_ctcrw_exp=get_final_response_tab(response_ctcrw_exp)
```

@@

The results are indeed similar to what we obtain using cubic splines with a low degree of freedom on both $\nu$ and $\tau$.

@@

post_coeff_re=post_coeff$coeff_re
post_coeff_fe=post_coeff$coeff_fe
post_coeff_lambda=exp(post_coeff$log_lambda)

@@

    H=array(rep(0.05^2*diag(2),n_post),dim=c(2,2,n_post))

@@

```{r propagated CI alpha log linear coefficients, message=FALSE}

@@

## Compute recovery distances

```{r}
make_recovery_distances_tab<-function(p) {

  parameter_names=names(mean_alpha)
  scaling=as.list(c(log(1-p),log(1+p)))
  names(scaling)=parameter_names
  estimates <- sapply(parameter_names,function(param) {
    mean_alpha[[param]]/scaling[[param]]})

  conf_intervals <- sapply(parameter_names, function(param) {
    sprintf("$[%.2f; %.2f]$", quant_alpha[[param]][1]/scaling[[param]], 
            quant_alpha[[param]][2]/scaling[[param]])
  })

  table_data <- data.frame(
    Parameter = 
      c("$D_{\\tau}^{ship}$","$D_{\\nu}^{ship}$"),
    Estimate = sprintf("$%.2f$", estimates),
    CI = conf_intervals,
    stringsAsFactors = FALSE
  )

  xtab <- xtable(
  table_data,
  caption = "Estimated recovery distances of the baseline values",
  label = paste0("table:recovery_distances_",p)
  )

  )
}

make_recovery_distances_tab(0.5)

make_recovery_distances_tab(0.1)
```

@@

We know want to investigate how the fitted model match the actual observations. To do this, we simulate trajectories from the fitted model before and during exposure. We ignore that the narwhals move in a restricted area and only look at the shape of the trajectories.
